HiC Contact Map Comaprison Using Graphlet Approach

### 2018-03-27 papers/manuscripts/manuscript.tex
updated manuscript

In this study, we plan to find dissimilarities between normal cells and cancerous cells,
through investigating HiC contact maps. 
We suspect that there are systematic differences between how chromosomes are structured
between normal cells and cancerous cells.
we 
Ideally, it is desirable to compare 3D structures of 
cell in order to make such comparisons.
However, the main challenge that we face is that 
3D structure of a cell is not readily available. Based on
\cite{adhikari2016chromosome3d}, fluorescence in situ hybridizaiton
(FISH) is used for investigating 3D configuration of chromosomes.
However, this method can only be used locally and cannot map
the whole structure of the chromosomes.
In orther to find dissimilarities in the 3D structure of 
chromosomes, we used HiC dataset.
The HiC method, which was developed by \cite{lieberman2009comprehensive}, captures interactions between 
chromosomal fragments in kilobase resolution. Based on HiC data, an
\textit{interaction frequency (IF) } matrix can be developed between \textit{loci} at a desired resolution.
A cell $IF_{ij}$ in an interaction frequency matrix captures the number of interaction detected
in HiC dataset between locus $i$ and locus $j$ in the genome.
An interaction matrix can be used to develop both inter- and intra-chromosomal interaction matrices.
We believe differences in interaction matrices can be found between normal cells and cancerous ones.
\end{abstract}
\section{Introduction}

@@

The authors of \cite{di2010fast} generalized the idea of graphlets to 
ordered graphs were the nodes are labeled in ascending order.
As can be viewed, there are a total of 14 orbits for graphlets of size
2 and 3 since the label of graphlets is also included in toplogy.
In the new definition, $d_v^i$ denotes the number of orbit $i$ touches 
node $v$. Each node, is then assigned a vector of length 14 
\footnote{number of orbits in graphlets of size 2 and 3}
$(d_v^1, d_v^2, ..., d_v^{14})$ 
and similarity of two nodes in two contact maps can be compared by
how geometrically close their corresponding vectors are.
\section{Materials and Methods}
\subsubsection{Notations}
In this paper, matrices and vectors are represented with bold
capital and bold small letters respectively.
matrix rows and columns are represented by a \textit{dot}
notation. For example, the $i$th row of matrix $M$ is
denoted by $M_{i.}$ and its $j$th column is represented
by $M_{.j}$.

@@

@@

The second measure of comparison that we use captures how 
similar two orbits are in terms of their count 
frequencies across loci between two contact maps. 
Each column in $S_{ij}$ can provide information
regarding the \textit{frequency distribution} of orbits throughout
the contact map $C_{ij}$. 
We can find how similar these distributions are to each other using
correlation measures.
These correlations are denoted by $\mathbf{m}^{\scriptscriptstyle A,B}_{i,j}$ and
can be calculate using
any plausible correlation measure. 
In this study, for each contact map,  we calculated
similarity between orbit distributions using Pearson's r 
correlation, which is computationally efficient.
However, pearson's r might not be able to capture
non-functional relationships between distributions. As a result, we
also used Maximal Information Coefficient (MIC) 
\cite{reshef2011detecting} in order to compare
correlations. MIC calculates mutual information (MI) between two
distributions, but utilizes dynamic programming in order adjust
bin sizes and numbers in order to achieve highest MI.
MIC values between two variables fall between 0 and 1,
with 0 meaning the two variables are completely independent
and 1 meaning one is dependant on the other.
We used both Pearson's r and MI in order to compare orbit
frequencies. Although results from both approaches were more
or less consistent, MIC showed higher robustness than Pearson's 
r method.

@@

\section{results}

Result of pair-wise contact map graphlet distances can is
illustrated in Figure \ref{fig:orbit-distances}.
Each point on the graph is the
average of the graphlet distance vector 
($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,j}$)
of the two cell lines specified in the legend.
 A one sided paired t-test
was conducted . The resulting p-values showed highly 
significant differeces from zero for all pairs of
cells. Detailed results of t-tests for each
contact map can be fount in supplementary materials.
For each contact map, each pair of cells are
order based on whether they are statistially
larger that the other or not. For example
for contact map $C_{i,j}$, the results 
of t-test is as follows:
\vspace{.2cm}\\
\small{
zero  $<$  call4-mit  $<$     all-rl  $<$     mit-rl  $<$   call4-rl  $<$    all-mit  $=$  all-call4
}
\vspace{.2cm}\\
which means that $\mathbf{d}^{\scriptscriptstyle CALL4, MIT}_{1,1}$ is statistically larger than 0,
but less than $\mathbf{d}^{\scriptscriptstyle ALL, RL}_{1,1}$ 
and so forth. We can also conclude that $\mathbf{d}^{\scriptscriptstyle ALL, CALL4}_{1,1}$ is not
statistically different from $\mathbf{d}^{\scriptscriptstyle ALL, MIT}_{1,1}$
between graphlets extracted from all contact maps. 
Refer to supplementary
material for results of all hypotheis tests.

\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_all.png}
        \caption{}
        \label{fig:orbit-distances_all}
    \end{subfigure}

    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_intra.png}
        \caption{}
        \label{fig:orbit-distances_intra}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_intra_separate.png}
        \caption{}
        \label{fig:orbit-distances_intra_separate}
    \end{subfigure}
    \caption{   Pair-wise graphlet signature difference for
                (\textbf{a}) all 276 contact maps
                ($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,j} \quad
                \forall i,j \in \{1 ... 23\} \quad \& \quad j \ge i$).
                (\textbf{b}) only the 23 intra-chromosomal contact maps
                ($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,i}$).
                (\textbf{c}) all 276 contact maps as well as corresponding
                standard errors.
                Each point on a graph is the result of averaging all
                the distances acroos all loci of that contact map.
                As can be viewd, although inter-chromosomal graphlets 
                do not show significant differences, RL's inter-chromosomal
                graphlets seem to be closer to MIT's. Also in some contact
                maps ALL graphlets show greater distance from MIT graphlets
                than the other two cell lines.
             }
    \label{fig:orbit-distances}
\end{figure*}

We caclulated pair-wise MIC for each orbit in each of the 276
contact maps from MIT data and ALL, RL and CALL4 data separately. 
Figure \ref{fig:contact_maps_correlations}
shows average orbit correlations across all contact maps, while
figure \ref{fig:orbits_correlations} 
demonstrates average correlations across all 72 orbits within each  contact map.
It is worth mentioning that \textit{interchromosomal} thresholded contact maps 
represent
a bipartide graph with the loci from each chromosome on one side. Due to this
bipartide nature of the graphs in inter-chromosomal maps,
count of certain orbits is always 0, resulting in
a correlation values of 0 for them as well.
We ignored these values  when we calculated averages across orbits 
in figure \ref{fig:contact_maps_correlations_all} since they
would result in a bias towards zero in averages. You can see the bias in 
figure \ref{fig:orbits_correlations_all} where average correlations of orbits
$\mathbb{Q} = \{3, 9, 10{\text -}14, 20{\text -}34, 39{\text -}48, 51{\text -}72\}$ 
are close to zero. In fact all correlations
corresponding to these orbits are 0 except for the ones between the same 
chromosomes which are illustrated in Figure 
\ref{fig:orbits_correlations_intra}.

\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/contact_maps_correlations_all.png}
        \caption{}
        \label{fig:contact_maps_correlations_all}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth]{figures/contact_maps_correlations_intra.png}
        \caption{}
        \label{fig:contact_maps_correlations_intra}
    \end{subfigure}
    \caption{pair-wise average contact map orbit correlations
             (a) for all contact maps($\bar{\mathbf{m}}^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,j} 
            \quad \forall i,j \in \{1 ... 23\} \quad \& \quad j \ge i$,
            average along the red \textit{vertical} arrow in figure 
            \ref{graphlet_correlation_schema})
             (b) only for intra-chromosomal contact maps
            ($\bar{\mathbf{m}}^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,i} 
            \quad \forall i \in \{1 ... 23\} $)
             These values are calculated by averaging over 
             pairwise correlations of all orbits in a 
             contact map.
             }
    \label{fig:contact_maps_correlations}
\end{figure*}

\begin{figure*}[t]
    \centering
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth, height=.2\paperheight]{figures/orbits_correlations_all.png}
        \caption{}
        \label{fig:orbits_correlations_all}
    \end{subfigure}
    \begin{subfigure}[b]{\textwidth}
        \includegraphics[width=\textwidth, height=.2\paperheight]{figures/orbits_correlations_intra.png}
        \caption{}
        \label{fig:orbits_correlations_intra}
    \end{subfigure}
    \caption{pair-wise average orbit correlations.
             In figure \ref{fig:orbits_correlations_all}, each point
             in the graph is the result of averaging pair-wise
             orbit correlations over all contact maps
             ($\frac{1}{276}\sum_{i=0}^{23}\sum_{j=i}^{23}{
             m^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,j, o}} \quad 
             \forall o \in \{0, 1, ..., 72\}$,
             average along the red \textit{horizontal} arrows in figure 
             \ref{graphlet_correlation_schema}), while
             each point in figure \ref{fig:orbits_correlations_intra}
             are averaged only over intra-chromosomal contact maps
             ($\frac{1}{23}\sum_{i=0}^{23}{
             m^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,i, o}} \quad 
         \forall o \in \{0, 1, ..., 72\}$).
             Counts for certain orbits are always zero in inter-chromosomal
             maps, leading to average value close to zero in 
             \ref{fig:orbits_correlations_all}.
             }
    \label{fig:orbits_correlations}
\end{figure*}
We have conducted one-sided
t-test in order to test whether the average correlations across
contact maps is equal to 1  and whether the average 
correlations across orbits is equal to 1. The results
for both test showed that all values are significantly less than 1.
Please refer to supplementary material for result of the t-tests.

\section{discussion}
\newpage
\section{Resources}
\textbf{Hi-C Datasets:}
\begin{enumerate}
    \item \href{https://github.com/rasoolianbehnam/watson}{Code base for this article}
    \item \href{http://sysbio.rnet.missouri.edu/T0510/tmp_download/link_to_download_genome_data/}
        {Datasets including cancerous cells}
    \item \href{https://bcm.app.box.com/v/aidenlab/folder/11234760671}{Original Datasets}
\end{enumerate}

>>>>

In this study, we investigated
dissimilarities between normal cells and cancerous cells,
through analyzing HiC contact maps. 
Our results show that certain orbit distributions
have significanly higher correlation 
between Leukemic cells.

\end{abstract}

\section{Introduction}

@@

The authors of \cite{di2010fast} generalized the idea of graphlets to 
ordered graphs were the nodes are labeled in ascending order.
As can be viewed, there are a total of 14 orbits for graphlets of size
2 and 3 since the label of graphlets is also included in toplogy.
In the new definition, $d_v^i$ denotes the number of orbit $i$ touches 
node $v$. Each node, is then assigned a vector of length 14 
\footnote{number of orbits in graphlets of size 2 and 3}
$(d_v^1, d_v^2, ..., d_v^{14})$ 
and similarity of two nodes in two contact maps can be compared by
how geometrically close their corresponding vectors are.

\subsubsection{Notations}
In this paper, matrices and vectors are represented with bold
capital and bold small letters respectively.
matrix rows and columns are represented by a \textit{dot}
notation. For example, the $i$th row of matrix $M$ is
denoted by $M_{i.}$ and its $j$th column is represented
by $M_{.j}$.

@@

\section{Materials and Methods}
We re-used Leukemic Hi-C libraries 
created in \cite{wang2013properties}
These libraries we sequenced 
for cases of primary human B-acute
lymphoblastic leukemia (B-ALL or ALL), 
the MHH-CALL-4 B-ALL cell
line (CALL4), 
and the follicular lymphoma cell-line (RL).
Just as \cite{wang2013properties}, 
we used normal B-cell line (GM068990)
from \cite{lieberman2009comprehensive} 
for our comparisons.
We created contact maps of resolution
500kb and normalized it using 
the \texttt{iced} package in python
developed by \cite{servant2015hic}.

@@

The second measure of comparison that we use captures how 
similar two orbits are in terms of their count 
frequencies across loci between two contact maps. 
Each column in $S_{ij}$ can provide information
regarding the \textit{frequency distribution} of orbits throughout
the contact map $C_{ij}$. 
We can find how similar these distributions are to each other using
correlation measures.
These correlations are denoted by $\mathbf{m}^{\scriptscriptstyle A,B}_{i,j}$ and
can be calculate using
any plausible correlation measure. 
In this study, for each contact map,  we calculated
similarity between orbit distributions using Pearson's r 
correlation, which is computationally efficient.
However, pearson's r might not be able to capture
non-functional relationships between distributions. As a result, we
also used Maximal Information Coefficient (MIC) 
\cite{reshef2011detecting} in order to compare
correlations. MIC calculates mutual information (MI) between two
distributions, but utilizes dynamic programming in order adjust
bin sizes and numbers in order to achieve highest MI.
MIC values between two variables fall between 0 and 1,
with 0 meaning the two variables are completely independent
and 1 meaning one is dependant on the other.
We used both Pearson's r and MIC in order to compare orbit
frequencies. Although results from both approaches were more
or less consistent, MIC showed higher robustness than Pearson's 
r method.

@@

\section{results and discussions}

Result of pair-wise contact map graphlet distances can is
illustrated in Figure \ref{fig:orbit-distances_all}.
Each point on the graph is the
average of the graphlet distance vector 
($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,j}$)
of the two cell lines specified in the legend.
 A one sided paired t-test
was conducted in order to compare distances to
zero and also to each other. 
The resulting p-values showed highly 
significant differeces from zero for all pairs of
cells. Detailed results of t-tests for each
contact map can be found in supplementary materials.
For each contact map, each pair of cells are
ordered based on whether they are statistially
larger that the other or not. For example
for contact map $C_{1,1}$, the results 
of t-test is as follows:
\vspace{.2cm}\\
\small{
zero  $<$  call4-mit  $<$     all-rl  $<$     mit-rl  $<$   call4-rl  $<$    all-mit  $=$  all-call4
}
\vspace{.2cm}\\
which means that 
$\mathbf{d}^{\scriptscriptstyle CALL4, MIT}_{1,1}$ 
is statistically larger than 0,
but less than $\mathbf{d}^{\scriptscriptstyle ALL, RL}_{1,1}$ 
and so forth. We can also conclude that 
$\mathbf{d}^{\scriptscriptstyle ALL, CALL4}_{1,1}$ is not
statistically different from 
$\mathbf{d}^{\scriptscriptstyle ALL, MIT}_{1,1}$
between graphlets extracted from all contact maps. 
Refer to supplementary
material for results of all hypotheis tests.

We caclulated pair-wise MIC for each orbit in each of the 276
contact maps from MIT data and ALL, RL and CALL4 data separately. 
Figure \ref{fig:contact_maps_correlations_all}
shows average orbit correlations across all contact maps, while
figure \ref{fig:orbits_correlations_all} 
demonstrates average correlations across all 72 orbits within each  contact map.
It is worth mentioning that interchromosomal thresholded contact maps 
represent
a bipartide graph with the loci from each chromosome on one side. Due to this
bipartide nature of the graphs in inter-chromosomal maps,
count of certain orbits is always 0, resulting in
a correlation values of 0 for them as well.
We ignored these values  when we calculated averages across orbits 
in figure \ref{fig:contact_maps_correlations_all} since they
would result in a bias towards zero in averages. You can see the bias in 
figure \ref{fig:orbits_correlations_all} where average correlations of orbits
$\mathbb{Q} = \{3, 9, 10{\text -}14, 20{\text -}34, 39{\text -}48, 51{\text -}72\}$ 
are close to zero. In fact all correlations
corresponding to these orbits are 0 except for the ones between the same 
chromosomes.

Figure \ref{fig:orbits_correlations_all} clearly illustrates certain
orbits of Leukemic have higher correlation to each other than to the
normal MIT cell. In fact our statistical analysis shows that 
\textit {for orbits in $\mathbb{Q}$, 
intra-leukemic orbit correlations are significantly higher
than leukemic-normal orbit correlations}. This implies
there are significant differences between normal and
leukemic cells in terms of their local structure.
\begin{figure}
    \centering
    \begin{subfigure}[b]{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/orbit-distances_all.png}
        \caption{}
        \label{fig:orbit-distances_all}
    \end{subfigure}
    \begin{subfigure}[b]{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/contact_maps_correlations_all.png}
        \caption{}
        \label{fig:contact_maps_correlations_all}
    \end{subfigure}
    \begin{subfigure}[b]{.5\textwidth}
        \includegraphics[width=\textwidth]{figures/orbits_correlations_all.png}
        \caption{}
        \label{fig:orbits_correlations_all}
    \end{subfigure}
    \caption{   
        \textbf{(a) Pair-wise graphlet signature difference for
        all 276 contact maps:}
        Each point on a graph is the result of averaging all
        the distances across all loci of that contact map.
        ($\bar{\mathbf{d}}^{\scriptscriptstyle A,B}_{i,j} \quad
        \forall i,j \in \{1 ... 23\} \quad \& \quad j \ge i$).  
        \vspace{.1cm} \\
        \textbf{(b) pair-wise average contact map orbit correlations
         for all contact maps:}
        ($\bar{\mathbf{m}}^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,j} 
        \quad \forall i,j \in \{1 ... 23\} \quad \& \quad j \ge i$:
        average along the red \textit{vertical} arrow in figure 
        \ref{graphlet_correlation_schema}).
        These values are calculated by averaging over 
        pairwise correlations of orbits of
        $\mathbb{Q}$ in a contact map.
        \vspace{.1cm} \\
        \textbf{(c) pair-wise average orbit correlations:}
        In figure \ref{fig:orbits_correlations_all}, each point
        in the graph is the result of averaging pair-wise
        orbit correlations over all contact maps
        ($\frac{1}{276}\sum_{i=0}^{23}\sum_{j=i}^{23}{
        m^{\scriptscriptstyle \scriptscriptstyle A,B}_{i,j, o}} \quad 
        \forall o \in \{0, 1, ..., 72\}$:
        average along the red \textit{horizontal} arrows in figure 
        \ref{graphlet_correlation_schema}).
        Counts for certain orbits are always zero in inter-chromosomal
        maps, leading to average value close to zero in 
        Figure \ref{fig:orbits_correlations_all}.
     }
    \label{fig:results_all}
\end{figure}

We have conducted one-sided
t-test in order to test whether the average correlations across
contact maps is equal to 1  and whether the average 
correlations across orbits is equal to 1. The results
for both test showed that all values are significantly less than 1.
Please refer to supplementary material for result of the 
full list of t-test results.

\newpage
\section{Resources}
\textbf{Hi-C Datasets:}
\begin{enumerate}
    \item \href{https://github.com/rasoolianbehnam/watson}{Code base for this article}
    \item \href{http://sysbio.rnet.missouri.edu/T0510/tmp_download/link_to_download_genome_data/}
        {Datasets including cancerous cells}
    \item \href{https://bcm.app.box.com/v/aidenlab/folder/11234760671}{Original Datasets}
\end{enumerate}
